### 2018-11-08 paper.tex
Initial commit

% MATH 5251 Final paper.

\documentclass[12pt]{article}

\usepackage{graphicx}
\usepackage{amssymb, amsmath, amsthm, amsfonts}
\usepackage{tikz}

\textwidth = 7 in
\textheight = 9.5 in
\oddsidemargin = -0.3 in
\evensidemargin = -0.3 in
\topmargin = -0.4 in
\headheight = 0.0 in
\headsep = 0.0 in
\parskip = 0.2in
\parindent = 0.0in

\newtheorem{thm}{Theorem}[section]
\newtheorem{cor}[thm]{Corollary}
\newtheorem{prop}{Proposition}
\newtheorem{defn}{Definition}[section]

\title{A Review of Data Compression Methods}
\author{Anonymous}

\begin{document}
\maketitle

In the age of ever-expanding quantities of digital data, the ability
to store this data efficiently is more important than ever. Throughout
the history of the field of computer science, various algorithms of
varying complexities and efficacies have been created in order to
reduce the amount of space required to store arbitrary strings of
binary data. This wide variety of methods are optimized for different
purposes: some algorithms attempt to reduce the amount of space
required to store raster images; some try to optimize audio storage;
some can optimize genetic data to an extremely high degree; and still
more attempt to work well for general-purpose use. Within the scope of
this paper, I will attempt to give an overview of a handful of popular
compression schemes, including general purpose as well domain-specific
methods, as well as a theoretical overview of the limits and potential
benefits of compression in general. This paper will only cover
lossless data encoding: that is, alogrithms which can reproduce a
perfect copy of the input data from its output. Lossy compression
schemes exist which can achieve incredible compression ratios, but
their methods typically include psychological ``tricks'' which allows
the algorithm to discard data which is less noticable to human
senses. As such, these methods are outside the scope of this paper.

\section{Theoretical Limits to Data Compression}

Data compression has many benefits as quickly reviewed above; however,
many limitations also exist. Note that for simplicity, we will limit
our discussion of compression schemes to binary schemes, as these are
the most wide spread in applications in the real world (i.e.,
compression of computer files.)

\begin{thm}
  For any given data compression scheme, there exist some input
  whose output is not shorter than its input. That is to say, there is
  no compression scheme which can compress every input bit-string.
\end{thm}

\begin{proof}
  Informally, we can describe a ``perfect compression scheme'' as some
  mapping which uniquely maps a given input bit-string to an output
  bit-string of strictly lesser length. More formally, a perfect
  compression scheme can be modeled as a function which maps input
  bit-strings to output bitstrings which are shorter. For any
  $n \in \mathbb{N}$, let $S_n$ be the set of bit-strings of length
  $n$, and let $T_n = \bigcup_{k = 1}^{n - 1} S_k$; that is, $T_n$ is
  the ``target'' set of bit-strings which are of lengths are strictly
  less than $n$. Suppose that there exists some ``perfect''
  compression scheme $C:S_n \to T_n$. The set $S_n$ has $2^n$ elements
  and $T_n$ has $\sum_{k=1}^{n-1}2^k = 2^n - 2$ elements. Since
  $2^n > 2^n - 2$, by the pigeonhole principle there must be at least
  one element in $T_n$ such that two distinct elements in $S_n$ map to
  this element through $C$. However, $C$ is defined to be a function,
  so this is a contradiction, and thus such a function $C$ can not
  exist.
\end{proof}

This result is not to say that compression is impossible; on the
contrary! It is very possible for a compression scheme to reduce
\emph{on average} the length of its inputs.

\section{Huffman Codes}

\begin{thm}
  This is a theorem.
\end{thm}

\begin{proof}
This is a proof.
\end{proof}

The equation $y=x$ is an in-line equation.

The equation
\[
y=x
\]
is a display style equation.

\section{Lempel-Ziv Based Methods}

If there's anything you don't know how to do with \LaTeX{}, try
googling it. If you still can't figure it out, email me.

\end{document}

### 2018-11-09 paper.tex
Add new stuff.

@@
 \usepackage{tikz}
+\usepackage[pdfusetitle]{hyperref}
 
@@
 
-\title{A Review of Data Compression Methods}
+\title{A Review of Lossless Data Compression Methods}
 \author{Anonymous}
@@
 this paper, I will attempt to give an overview of a handful of popular
-compression schemes, including general purpose as well domain-specific
-methods, as well as a theoretical overview of the limits and potential
-benefits of compression in general. This paper will only cover
-lossless data encoding: that is, alogrithms which can reproduce a
-perfect copy of the input data from its output. Lossy compression
-schemes exist which can achieve incredible compression ratios, but
-their methods typically include psychological ``tricks'' which allows
-the algorithm to discard data which is less noticable to human
-senses. As such, these methods are outside the scope of this paper.
+compression schemes, including general purpose in addition to
+domain-specific methods, as well as a theoretical overview of the
+limits and potential benefits of compression in general. This paper
+will only cover lossless data encoding: that is, alogrithms which can
+reproduce a perfect copy of the input data from its output. Lossy
+compression schemes exist which can achieve incredible compression
+ratios, but their methods typically include psychological ``tricks''
+which allows the algorithm to discard data which is less noticable to
+human senses. As such, these methods are outside the scope of this
+paper.
 
@@
 the most wide spread in applications in the real world (i.e.,
-compression of computer files.)
+compression of computer files.) The first limitation we shall discuss
+is the impossibility of a universally applicable, or ``perfect,''
+compression scheme.
 
 \begin{thm}
-  For any given data compression scheme, there exist some input
-  whose output is not shorter than its input. That is to say, there is
-  no compression scheme which can compress every input bit-string.
+  For any given data compression scheme, there exists at least one
+  arbitrary input data whose ``compressed'' form is not strictly
+  shorter than it. That is to say, there is no compression scheme
+  which can compress every potential input.
 \end{thm}
@@
   Informally, we can describe a ``perfect compression scheme'' as some
-  mapping which uniquely maps a given input bit-string to an output
-  bit-string of strictly lesser length. More formally, a perfect
-  compression scheme can be modeled as a function which maps input
-  bit-strings to output bitstrings which are shorter. For any
+  mapping which uniquely maps given input bit-strings to output
+  bit-strings which are shorter. More formally, a perfect compression
+  scheme can be modeled as a bijection which maps input bit-strings to
+  output bit-strings of strictly lesser length. For any
   $n \in \mathbb{N}$, let $S_n$ be the set of bit-strings of length
   $n$, and let $T_n = \bigcup_{k = 1}^{n - 1} S_k$; that is, $T_n$ is
-  the ``target'' set of bit-strings which are of lengths are strictly
-  less than $n$. Suppose that there exists some ``perfect''
-  compression scheme $C:S_n \to T_n$. The set $S_n$ has $2^n$ elements
-  and $T_n$ has $\sum_{k=1}^{n-1}2^k = 2^n - 2$ elements. Since
-  $2^n > 2^n - 2$, by the pigeonhole principle there must be at least
-  one element in $T_n$ such that two distinct elements in $S_n$ map to
-  this element through $C$. However, $C$ is defined to be a function,
-  so this is a contradiction, and thus such a function $C$ can not
-  exist.
+  the ``target'' set of bit-strings which are of lengths strictly less
+  than $n$. Suppose that there exists some perfect compression scheme
+  $C:S_n \to T_n$. The set $S_n$ has $2^n$ elements and $T_n$ has
+  $\sum_{k=1}^{n-1}2^k = 2^n - 2$ elements. Since
+  $|S_n| = 2^n > 2^n - 2 = |T_n|$, by the pigeonhole principle there
+  must be at least one element $t \in T_n$ such that at least two
+  distinct elements in $S_n$ map to $t$ through $C$. However, $C$ is a
+  bijection by definition, so this is a contradiction. Thus a perfect
+  compression algorithm cannot exist for arbitrary-length bit-strings.
 \end{proof}
 
-This result is not to say that compression is impossible; on the
-contrary! It is very possible for a compression scheme to reduce
-\emph{on average} the length of its inputs.
+A brief note on the implications of this result: this applies to all
+lossless compression schemes, as a compression scheme must be able to
+uniquely decompress a given compressed bit-string. If it cannot do
+this, the compression scheme is in effect discarding the data of at
+least one of its potential inputs and is then clearly not
+``lossless.''
 
-\section{Huffman Codes}
+This result also raises the question of which inputs are more likely
+to be compressable in general.  (That is, regardless of compression
+algorithm implementation, what sorts of input can yield a higher
+compression ratio?) In practice, most compression algorithms function
+by finding common repeated patterns within the input data, and then
+replacing each instance of this input with a shorter ``prefix code.''
+Intuitively, this suggests that data with a regular structure is more
+amenable to compression than more random, unpredictable data.
 
 \begin{thm}
-  This is a theorem.
+  The average word length of an optimal prefix code for a given source
+  is greater than or equal to the entropy of that source.
 \end{thm}
@@
 \begin{proof}
-This is a proof.
+  A proof of this theorem can be found in \cite{Shannon}.
 \end{proof}
 
-The equation $y=x$ is an in-line equation.
-
-The equation
-\[
-y=x
-\]
-is a display style equation.
+This result places a lower bound on the \emph{efficiency} of prefix
+codes in general, and suggests the sorts of data which are less
+compressable, and even uncompressable (since such an input must
+exist!) Since entropy of some channel $X$ with an alphabet $C$ is
+defined to be $H(X) = -\sum_{c \in C} P(c)\log_2{c}$, we can
+illustrate the implications of this result with a simple
+example. Clearly for a channel using binary alphabet (i.e.,
+$C = \{0, 1\}$) with a uniform distribution of 1s and 0s, the entropy
+of this channel is $H(X) = -2 * \frac{1}{2}\log_2{\frac{1}{2}} = 1$,
+so some optimal prefix code must have an average word length greater
+than or equal to 1; obviously, the simplest prefix code to encode
+$\{0, 1\}$ is $\{0, 1\}$ itself. Thus, a this channel is
+incompressible.
 
@@
 
-If there's anything you don't know how to do with \LaTeX{}, try
-googling it. If you still can't figure it out, email me.
+The Lempel-Ziv compression algorithm, or simply LZ for short, is one
+of the most widespread compression schemes in use, when including its
+large family of derivative methods. It was first introduced in 1977 in
+\cite{LempelZiv}.
+
+\section{The Free Lossless Audio Codec}
+
+Since knowing the structure of a given input can allow the
+construction of a more efficient data compression scheme, it is
+worthwhile to study some compression scheme which is specifically
+designed to store a particular type of data. In this section I will
+describe the structure and function of the Free Lossless Audio Codec,
+known more commonly by its acronym FLAC, a popular format for
+high-fidelity audio storage.
+
+\bibliography{biblio}{}
+\bibliographystyle{plain}
 

### 2018-11-15 paper.tex
Progress on paper lol

@@
 \usepackage[pdfusetitle]{hyperref}
+\usepackage{apacite}
 
@@
 illustrate the implications of this result with a simple
-example. Clearly for a channel using binary alphabet (i.e.,
-$C = \{0, 1\}$) with a uniform distribution of 1s and 0s, the entropy
-of this channel is $H(X) = -2 * \frac{1}{2}\log_2{\frac{1}{2}} = 1$,
-so some optimal prefix code must have an average word length greater
-than or equal to 1; obviously, the simplest prefix code to encode
-$\{0, 1\}$ is $\{0, 1\}$ itself. Thus, a this channel is
-incompressible.
+example. Suppose we are attempting to compress a particular channel
+$X$ which is text document in which the frequencies of every letter of
+the Latin alphabet are distributed equally. (It is extremely unlikely
+that a text in any human language would have this property; it is used
+here to illustrate a pathological edge case in general.) Then the
+average word length must be greater than
+$26 * -\frac{1}{26} \log_2{\frac{1}{26}} \approx 4.7$ bits. If we were
+to construct a naive code to encode the Latin alphabet, for instance
+$A=1, B=2,...$, we would require 5 bits for each letter. Shannon's
+Noiseless Coding Theorem says this is more or less as good as we can
+do, so for this case it is very difficult to actually compress our
+input. Consider another document in which case one letter, let's say
+$E$, occurs significantly more often than the others, for instance
+$\frac{27}{52}$ of the time, relative to $\frac{1}{52}$ for every
+other letter. Then by Shannon's Noiseless Coding Theorem, it is
+possible to construct a code with an average word length greater than
+$25 * -\frac{1}{52} \log_2{\frac{1}{52}} - \frac{27}{52}
+\log_2{\frac{27}{25}} \approx 3.2$, meaning that we can compress such
+this document approximately $33\%$ more efficiently than the uniform
+one. This situation is somewhat contrived, but it serves to illustrate
+the fundamental limits of prefix codes in general.
 
@@
 
-The Lempel-Ziv compression algorithm, or simply LZ for short, is one
-of the most widespread compression schemes in use, when including its
-large family of derivative methods. It was first introduced in 1977 in
-\cite{LempelZiv}.
+The Lempel-Ziv compression algorithm, commonly referred to as LZ for
+short, is one of the most widespread compression schemes in use,
+especially when including its large family of derivative methods. It
+was first introduced in 1977. \cite{LempelZiv} Algorithms based on
+the Lempel-Ziv algorithms are very efficient and work well with a wide
+variety of inputs.
 
@@
 
-\bibliography{biblio}{}
-\bibliographystyle{plain}
+\bibliographystyle{apacite}
+\bibliography{biblio}
 

### 2018-11-16 paper.tex
Add first information about LZW.

@@
 \title{A Review of Lossless Data Compression Methods}
-\author{Anonymous}
+\author{Blargh}
 
@@
 $26 * -\frac{1}{26} \log_2{\frac{1}{26}} \approx 4.7$ bits. If we were
-to construct a naive code to encode the Latin alphabet, for instance
+to construct a naïve code to encode the Latin alphabet, for instance
 $A=1, B=2,...$, we would require 5 bits for each letter. Shannon's
@@
 other letter. Then by Shannon's Noiseless Coding Theorem, it is
-possible to construct a code with an average word length greater than
+possible to construct a prefix code with an average word length
+greater than
 $25 * -\frac{1}{52} \log_2{\frac{1}{52}} - \frac{27}{52}
@@
 
+Prefix codes on their own are extremely useful in communication
+applications, but their true power is revealed when used as a building
+block within more powerful, sophisticated compression schemes.
+
 \section{Lempel-Ziv Based Methods}
 
-The Lempel-Ziv compression algorithm, commonly referred to as LZ for
-short, is one of the most widespread compression schemes in use,
-especially when including its large family of derivative methods. It
-was first introduced in 1977. \cite{LempelZiv} Algorithms based on
-the Lempel-Ziv algorithms are very efficient and work well with a wide
-variety of inputs.
+The Lempel-Ziv compression algorithm, commonly referred to as LZ77 or
+LZ78 for short, is one of the most widespread compression schemes in
+use, especially when including its large family of derivative
+methods. It was first introduced in 1977. \cite{LempelZiv} Algorithms
+based on the Lempel-Ziv algorithms are relatively efficient and work
+well with a wide variety of inputs.
+
+The algorithm is based on a dictionary-style approach; that is to say,
+it creates a dictionary of codewords within the input bit-string
+as the algorithm progresses.
+
+\subsection{Lempel-Ziv-Welch}
+
+The Lempel-Ziv-Welch compression algorithm was an extension of the
+Lempel-Ziv algorithm which was introduced soon after the publication
+of the original LZ77. \cite{LempelZivWelch}
 
@@
 worthwhile to study some compression scheme which is specifically
-designed to store a particular type of data. In this section I will
+designed to store a particular type of data. In this secion I will
 describe the structure and function of the Free Lossless Audio Codec,

### 2018-11-17 paper.tex
Update with DEFLATE RFC.

@@
 The Lempel-Ziv compression algorithm, commonly referred to as LZ77 or
-LZ78 for short, is one of the most widespread compression schemes in
-use, especially when including its large family of derivative
-methods. It was first introduced in 1977. \cite{LempelZiv} Algorithms
-based on the Lempel-Ziv algorithms are relatively efficient and work
-well with a wide variety of inputs.
+LZ78, or even simply LZ for short, is one of the most widespread
+compression schemes in use, especially when including its large family
+of derivative methods. It was first introduced in
+1977. \cite{LempelZiv} Algorithms based on the Lempel-Ziv algorithms
+are relatively efficient and work well with a wide variety of inputs.
 
 The algorithm is based on a dictionary-style approach; that is to say,
-it creates a dictionary of codewords within the input bit-string
-as the algorithm progresses.
+it creates a dictionary of codewords within the input data as the
+algorithm progresses. Lempel-Ziv uses a ``sliding-window'' technique
+to construct its dictionary. For each set of bytes it attempts to
+compress, it inspects its sliding window, which extends some set
+amount behind the bytes under consideration. In this way, the
+algorithm builds a prefix code as it proceeds through the file. A
+``prefix'' can be one or more characters long. For each point in the
+input data, the algorithm consults the prefix table of prefixes
+contained in the sliding window, which is a set amount of data prior
+to the current point. If the current prefix is found in the prefix
+table, the algorithm outputs a data triplet indicating how far back in
+the window the prefix occurred, the length of the prefix, and finally
+the character which comes after the prefix. This means that the very
+first characters in the input data will be encoded as 1-character
+prefixes. (For our very first characters in the input data, since
+there can't be any prefixes in the sliding window yet, the encoded
+form will consist of zero-length prefixes followed by the input
+character literals.)
+
+The basic method introduced in Lempel and Ziv's original paper is
+vulnerable to some significant drawbacks. Since each individual byte
+in the input can take up at most three bytes in the output, it is
+possible that some pathological input data can have a compressed form
+which is substantially larger, especially if the sliding window is
+small. As such, this suggests a good jumping-off point for creating
+improvements upon the original algorithm.
 
@@
 Lempel-Ziv algorithm which was introduced soon after the publication
-of the original LZ77. \cite{LempelZivWelch}
+of the original LZ77. \cite{LempelZivWelch} The LZW variant is
+extremely widespread in applications; for instance, one of its many
+uses is as part of the GIF image format.
+
+\subsection{DEFLATE}
+
+Another algorithm based on the Lempel-Ziv foundation is the DEFLATE
+algorithm, codified in RFC 1951. \cite{RFC1951}
 

### 2018-11-17 paper.tex
Add internal references.

@@
 \usepackage{amssymb, amsmath, amsthm, amsfonts}
+\usepackage{thmtools}
 \usepackage{tikz}
 \usepackage[pdfusetitle]{hyperref}
+\usepackage{cleveref}
 \usepackage{apacite}
@@
 
-\newtheorem{thm}{Theorem}[section]
-\newtheorem{cor}[thm]{Corollary}
-\newtheorem{prop}{Proposition}
-\newtheorem{defn}{Definition}[section]
+\declaretheorem[numberwithin=section]{Theorem}
 
 \title{A Review of Lossless Data Compression Methods}
-\author{Blargh}
+\author{Citoyen}
 
@@
 
-\begin{thm}
+\begin{Theorem}
+  \label{imperfect-compression}
   For any given data compression scheme, there exists at least one
@@
   which can compress every potential input.
-\end{thm}
+\end{Theorem}
 
@@
 
-\begin{thm}
+\begin{Theorem}
+  \label{snct}
   The average word length of an optimal prefix code for a given source
   is greater than or equal to the entropy of that source.
-\end{thm}
+\end{Theorem}
 
@@
 The basic method introduced in Lempel and Ziv's original paper is
-vulnerable to some significant drawbacks. Since each individual byte
-in the input can take up at most three bytes in the output, it is
-possible that some pathological input data can have a compressed form
-which is substantially larger, especially if the sliding window is
-small. As such, this suggests a good jumping-off point for creating
-improvements upon the original algorithm.
+vulnerable to some significant drawbacks. For instance, since each
+individual byte in the input can take up at most three bytes in the
+output, it is possible that some pathological input data can have a
+compressed form which is substantially larger, especially if the
+sliding window is small. (Clearly, a highly entropic input is much
+more likely to have fewer recurring prefixes within any given sliding
+window, which is consistent with our result from
+\Cref{imperfect-compression}) As such, this suggests a good jumping-off
+point for creating improvements upon the original algorithm.
 

### 2018-11-19 paper.tex
Updates...
Finish discussion of LZW.

@@
 \title{A Review of Lossless Data Compression Methods}
-\author{Citoyen}
+\author{someone idk}
 
@@
 more likely to have fewer recurring prefixes within any given sliding
-window, which is consistent with our result from
-\Cref{imperfect-compression}) As such, this suggests a good jumping-off
-point for creating improvements upon the original algorithm.
+window. Thus a prefix code based on such an output would have a higher
+average word length, which is consistent with our result from
+\Cref{snct}.) Another drawback of the LZ method is its performance;
+searching the sliding window for matches, as well as updating the
+window prefix list, can be an expensive operation. As such, these
+drawbacks suggest a good jumping-off point for creating improvements
+upon the original algorithm.
 
@@
 
-The Lempel-Ziv-Welch compression algorithm was an extension of the
+One such improvement was the Lempel-Ziv-Welch compression algorithm,
+commonly referred to as LZW. It was created as an extension of the
 Lempel-Ziv algorithm which was introduced soon after the publication
-of the original LZ77. \cite{LempelZivWelch} The LZW variant is
-extremely widespread in applications; for instance, one of its many
-uses is as part of the GIF image format.
+of the original LZ77. The LZW variant is extremely widespread in
+applications; for instance, one of its many uses is as part of the GIF
+image format. LZW's improvements upon the LZ algorithm are fairly
+straightforward, but they allow for significant improvements to the
+simplicity of implementation. According to the author, the LZW
+algorithm can take operate at a speed of three clock cycles (CPU
+instructions) per input symbol. \cite{LempelZivWelch}
+
+LZW is able to achieve this speed by using a single symbol table for
+the entire input, adaptively building up as the algorithm processes
+it. Instead of using a sliding window, LZW instead begins by
+initializing a string table with every single-character prefix. As it
+parses the input, it uses a greedy algorithm to match strings from the
+input to the strings already seen in the table. It does this by
+checking if some prefix in the input has already been seen in the
+table; if so, it appends the immediately next character from the input
+to the prefix and queries the string table again. If, on the other
+hand, the appended prefix is \emph{not} found, the algorithm outputs
+the code of the prefix \emph{without} the next character appended (the
+code for which is retrieved from the string table) and then adding the
+appended prefix to the string table. The prefix is set to the last
+character of the prefix and the process repeats.
+
+As we can see, the entire algorithm can be succinctly summarized in a
+few sentences and can also be (hopefully) easily understood. This
+characteristic makes it useful for implementers, and also has benefits
+in terms of performance. Since it uses a very simple process for
+encoding, it is able to greatly reduce the number of clock cycles
+needed for encoding or decoding each character. In addition, it is
+also simple to implement performance-enhancing features, such as using
+a tree data structure with hashing to store string prefixes, buffering
+the input data to conserve memory, and so forth. Drawbacks of the LZW
+method include lower compression ratios than other, more advanced
+methods. For instance, the author claims a ratio of 1.8 on English
+text, compared to a 2.4 ratio achieved by more sophisticated
+algorithms. \cite{LempelZivWelch}
+
+The LZ family of compression codes is massive, and LZW is just one
+single member. Another, possibly even more widespread variant is the
+DEFLATE algorithm.
 
@@
 Another algorithm based on the Lempel-Ziv foundation is the DEFLATE
-algorithm, codified in RFC 1951. \cite{RFC1951}
+format, codified in RFC 1951. \cite{RFC1951} DEFLATE is not
+``standardized'' by any recognized governing body, though it is very
+widely used and has become a de facto standard by virtue of its use in
+the \texttt{gzip} compression utility, as well as in the ubiquitous
+\texttt{zlib} library. RFC 1951 specifies the technical minutia of the
+implementation of the file format, as well as its algorithm itself. As
+such, it is not useful to describe the full file format within this
+paper, only the algorithm will be outlined.
 
@@
 
+\section{Conclusion}
+
 \bibliographystyle{apacite}

### 2018-11-19 paper.tex
Add DEFLATE.

@@
 commonly referred to as LZW. It was created as an extension of the
-Lempel-Ziv algorithm which was introduced soon after the publication
-of the original LZ77. The LZW variant is extremely widespread in
+Lempel-Ziv algorithm and was introduced soon after the publication of
+the original LZ77. The LZW variant is extremely widespread in
 applications; for instance, one of its many uses is as part of the GIF
@@
 The LZ family of compression codes is massive, and LZW is just one
-single member. Another, possibly even more widespread variant is the
-DEFLATE algorithm.
+single member. Possibly even more widespread is the DEFLATE algorithm,
+used in many compressed file formats.
 
@@
 paper, only the algorithm will be outlined.
+
+In general, compression proceeds as follows: for a given block of data
+withint the input, the algorithm constructs a Huffman prefix coding
+for the input alphabet. Information about the prefix coding is then
+stored within the output block header, which contains metadata about the
+compressed output. DEFLATE supports ``fixed'' and ``dynamic'' Huffman
+codes. Fixed Huffman codes are defined in the DEFLATE specification
+and can be simply hard-coded in the compressor/decompressor
+software. Dynamic Huffman codes, on the other hand, are created by the
+compressor and their information is serialized and stored in the block
+header. The compressed data is then run through an implementation of
+the LZ77 algorithm from earlier.
+
+DEFLATE allows us to reap the benefits both of the the LZ77 algorithm
+as well as the more basic Huffman prefix coding scheme, all while
+using a more-or-less standardized file format for easy interchange.
 

### 2018-11-19 paper.tex
Add basic FLAC information.

@@
 known more commonly by its acronym FLAC, a popular format for
-high-fidelity audio storage.
+high-fidelity audio storage. \cite{flac}
+
+Since FLAC is designed exclusively for storing high-fidelity audio
+data, it is able to exploit characteristics specific to sound in order
+to achieve a better compression ratio than a simple general purpose
+compression scheme would be able to. For instance, FLAC's designers
+realize that stereo sound streams often correlate very closely. As
+such, FLAC allows the compressor to encode input in multiple ways;
+left and right can be encoded independently, or as ``mid'' and
+``side'' channels, where the mid channel is an average of the left and
+right channels and the side is the difference of the two. For highly
+correlated left/right channels, this encoding can result in
+significant lossless compression gains.
+
+The primary means of compression, however, is through model/residual
+coding. For each block of raw input, the FLAC compressor tries to fit
+the signal to a parameterized function, such as a polynomial for
+instance, and then stores the function. It then computes the
+residuals, or the differences between the raw audio signal and the
+parameterized function at each sample. This residual data has a
+particular distribution which is efficiently stored using a specific
+form of Huffman codes, called Rice codes. \cite{flac}
 

### 2018-11-19 paper.tex
Let's just say this is an ok first draft. probably has problems.

@@
 \title{A Review of Lossless Data Compression Methods}
-\author{someone idk}
+\author{5251 Student}
 
@@
 to store this data efficiently is more important than ever. Throughout
-the history of the field of computer science, various algorithms of
-varying complexities and efficacies have been created in order to
-reduce the amount of space required to store arbitrary strings of
-binary data. This wide variety of methods are optimized for different
+the history of the field of computer science, algorithms of varying
+complexities and efficacies have been created in order to reduce the
+amount of space required to store arbitrary strings of binary
+data. This wide variety of methods are optimized for different
 purposes: some algorithms attempt to reduce the amount of space
@@
 
+The FLAC specification is of course much more detailed than this
+high-level overview, but a brief description of FLAC's methods is more
+than sufficient to demonstrate how exploiting characteristics of a
+particular data domain can lead to huge increases in compression
+efficiency.
+
 \section{Conclusion}
+
+Data compression methods can be used to reduce the amount of space
+required to store digital data. However, tradeoffs have to be made
+when compressing since it is impossible for an algorithm to be able to
+compress any arbitrary input; some input must result in an output of
+equal length. In addition, the compressability of an input is highly
+dependent on its entropy. Many different compression algorithms exist,
+the most common of which belong to the Lempel-Ziv family. Finally,
+many domain-specific algorithms exist to gain extra efficiency when
+applied to their particular domain. FLAC is one such algorithm which
+is used for lossless audio storage and transmission.
 

### 2018-12-05 presentation/presentation.tex
Add first part of presentation.

\documentclass[mathserif,notheorems]{beamer} % option notheorems
\usepackage{amssymb, amsmath, amsthm, amsfonts}
\usepackage{cleveref}
\usepackage{thmtools}
\usetheme{Berlin}
\title{A Review of Lossless Data Compression Algorithms}
\author{Jack Stanek}

\declaretheorem[numberwithin=section]{Theorem}

\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\begin{frame}
  \tableofcontents
\end{frame}

\begin{frame}
  \section{A Brief Introduction to Data Compression}
  \subsection{Informal Overview}
  \frametitle{What is data compression?}

  A data compression scheme refers to any system for reducing the
  number of symbols required to represent arbitrary information.

  Some commonplace (possibly lossy!) examples:

  \begin{itemize}
    \item \texttt{*.mp3} files
    \item \texttt{*.zip} files
    \item Abbreviations in writing
  \end{itemize}

  Clearly, this list is not exhaustive, but it provides an intuition
  to how these systems function.
\end{frame}

\begin{frame}
  \subsection{Theoretical Background}
  Information can be represented as a series of symbols
  \begin{itemize}
    \item Text is a string of letters in langauge's alphabet
    \item Numbers are strings of Arabic numerals
  \end{itemize}
  For some alphabet $\Sigma$, the set of all compression schemes
  $C_\Sigma$ is the set of bijections
  $c : \Sigma^n \to \bigcup_{k = 1}^{n} \Sigma^k$ for all
  $n \in \mathbb{N}$ such that $n \geq 2$. A compression scheme is
  considered ``perfect'' if and only if its image is the set of all
  strings of length strictly less than that of the domain.
\end{frame}

\begin{frame}
  \subsection{Impossibility of perfect compression}
  \begin{Theorem}
    No perfect compression scheme exists.
  \end{Theorem}
\end{frame}

\begin{frame}

\end{frame}

\end{document}

### 2018-12-08 paper.tex
Updates

@@
 \usepackage{amssymb, amsmath, amsthm, amsfonts}
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
 \usepackage{thmtools}
@@
 \title{A Review of Lossless Data Compression Methods}
-\author{5251 Student}
+\author{Jack Stanek}
 
@@
 which allows the algorithm to discard data which is less noticable to
-human senses. As such, these methods are outside the scope of this
-paper.
+human senses. As this is a mathematics paper and not a psychology
+paper, these methods are outside the scope of this document.
 
@@
 our discussion of compression schemes to binary schemes, as these are
-the most wide spread in applications in the real world (i.e.,
+the most widespread in applications in the real world (i.e.,
 compression of computer files.) The first limitation we shall discuss
@@
 
-The Lempel-Ziv compression algorithm, commonly referred to as LZ77 or
-LZ78, or even simply LZ for short, is one of the most widespread
-compression schemes in use, especially when including its large family
-of derivative methods. It was first introduced in
-1977. \cite{LempelZiv} Algorithms based on the Lempel-Ziv algorithms
-are relatively efficient and work well with a wide variety of inputs.
+The Lempel-Ziv compression algorithm, introduced in two variations in
+1977 and 1978, is one of the most widespread compression schemes in
+use, especially when including its large family of derivative
+methods. Algorithms based on the Lempel-Ziv algorithms are relatively
+efficient and work well with a wide variety of inputs.
 
@@
 it creates a dictionary of codewords within the input data as the
-algorithm progresses. Lempel-Ziv uses a ``sliding-window'' technique
-to construct its dictionary. For each set of bytes it attempts to
-compress, it inspects its sliding window, which extends some set
-amount behind the bytes under consideration. In this way, the
-algorithm builds a prefix code as it proceeds through the file. A
-``prefix'' can be one or more characters long. For each point in the
-input data, the algorithm consults the prefix table of prefixes
-contained in the sliding window, which is a set amount of data prior
-to the current point. If the current prefix is found in the prefix
-table, the algorithm outputs a data triplet indicating how far back in
-the window the prefix occurred, the length of the prefix, and finally
-the character which comes after the prefix. This means that the very
-first characters in the input data will be encoded as 1-character
-prefixes. (For our very first characters in the input data, since
-there can't be any prefixes in the sliding window yet, the encoded
-form will consist of zero-length prefixes followed by the input
-character literals.)
+algorithm progresses. The LZ77 variant uses a ``sliding-window''
+technique to construct its dictionary as it compresses. \cite{lz77}
+For each set of bytes it attempts to compress, it inspects its sliding
+window, which extends some set amount behind the bytes under
+consideration. In this way, the algorithm builds a prefix code as it
+proceeds through the file. A ``prefix'' can be one or more characters
+long. For each point in the input data, the algorithm consults the
+prefix table of prefixes contained in the sliding window, which is a
+set amount of data prior to the current point. If the current prefix
+is found in the prefix table, the algorithm outputs a 3-tuple
+indicating how far back in the window the prefix occurred, the length
+of the prefix, and finally the character which comes after the
+prefix. This means that the very first characters in the input data
+will be encoded as 1-character prefixes. (For our very first
+characters in the input data, since there can't be any prefixes in the
+sliding window yet, the encoded form will consist of zero-length
+prefixes followed by the input character literals.)
+
+To illustrate: suppose we are to encode the string ``oodles of
+noodles'' using LZ77, with a window of length 8. The string is 16
+characters long, requiring 16 bytes to store. The current character is
+set to ``o'', the first character of the string. However, since there
+are two ``o'' characters in a row, the algorithm can ``overshoot'' the
+end of the window in order to efficiently implement run-length
+encoding (that is, recording a character and the number of copies
+folloiwng it.) Thus, \texttt{(2, 0, 'o')} is emitted, and the current
+character is incremented by two. The algorithm then emits
+single-character prefix 3-tuples for \texttt{d, l, e, s, ' ',} (the
+space character), \texttt{o}, and \texttt{f}. At
+
+The LZ78 algorithm is functionally equivalent to the LZ77 algorithm,
+but it forgoes the sliding window altogether, and instead uses a
+``linked dictionary''. \cite{lz78} As the compressor processes the
+input data, it maintains a matching index to longest prefix already
+stored in the dictionary. If the prefix created by appending the
+current character to the last matched prefix is not found in the
+dictionary, the algorithm stores the character at the next available
+index slot along with a pointer to the index of the original prefix,
+and outputs the index of the longest matching prefix followed by the
+character. Algorithmically, this method is equivalent to LZ77.
 
@@
 the code of the prefix \emph{without} the next character appended (the
-code for which is retrieved from the string table) and then adding the
-appended prefix to the string table. The prefix is set to the last
-character of the prefix and the process repeats.
+code for which is retrieved from the string table) and then adds the
+appended prefix to the string table. The current prefix is finally set
+to the last character of the prefix and the process repeats.
 
@@
 
-Another algorithm based on the Lempel-Ziv foundation is the DEFLATE
-format, codified in RFC 1951. \cite{RFC1951} DEFLATE is not
-``standardized'' by any recognized governing body, though it is very
-widely used and has become a de facto standard by virtue of its use in
-the \texttt{gzip} compression utility, as well as in the ubiquitous
-\texttt{zlib} library. RFC 1951 specifies the technical minutia of the
-implementation of the file format, as well as its algorithm itself. As
-such, it is not useful to describe the full file format within this
-paper, only the algorithm will be outlined.
+An exmaple of an efficient implementation of a compression algorithm
+can be seen in the DEFLATE format, codified in the RFC 1951
+memo. \cite{RFC1951} DEFLATE is not ``standardized'' by any recognized
+governing body, though it is very widely used and has become a de
+facto standard by virtue of its use in the \texttt{gzip} compression
+utility, as well as in the ubiquitous \texttt{zlib} library. RFC 1951
+specifies the technical minutia of the implementation of the file
+format, as well as its algorithm itself. As such, it is not useful to
+describe the full file format within this paper, and so only the
+algorithm will be outlined.
 
-In general, compression proceeds as follows: for a given block of data
-withint the input, the algorithm constructs a Huffman prefix coding
-for the input alphabet. Information about the prefix coding is then
-stored within the output block header, which contains metadata about the
+Briefly, compression proceeds as follows: for a given block of data
+within the input, the algorithm constructs a Huffman prefix coding for
+the input alphabet. Information about the prefix coding is then stored
+within the output block header, which contains metadata about the
 compressed output. DEFLATE supports ``fixed'' and ``dynamic'' Huffman
@@
 header. The compressed data is then run through an implementation of
-the LZ77 algorithm from earlier.
+the LZ77 algorithm. As such, this combination of prefix coding with a
+powerful compression scheme allows compression tools to squeeze every
+last bit of possible compression out of their data.
 
@@
 coding. For each block of raw input, the FLAC compressor tries to fit
-the signal to a parameterized function, such as a polynomial for
-instance, and then stores the function. It then computes the
-residuals, or the differences between the raw audio signal and the
-parameterized function at each sample. This residual data has a
-particular distribution which is efficiently stored using a specific
-form of Huffman codes, called Rice codes. \cite{flac}
+the signal to a parameterized function, typically a polynomial, and
+then stores the function. It then computes the residuals, or the
+differences between the raw audio signal and the parameterized
+function at each sample. This residual data has a particular
+distribution which is efficiently stored using a specific form of
+Huffman codes, called Rice codes. \cite{flac}
 

### 2018-12-08 presentation/presentation.tex
Updates

@@
 
-  A data compression scheme refers to any system for reducing the
-  number of symbols required to represent arbitrary information.
+  Informally, a data compression scheme is some method for reducing
+  the amount of space required to store a given amount of data.
 
-  Some commonplace (possibly lossy!) examples:
+  Some commonplace examples:
 
@@
 
-  Clearly, this list is not exhaustive, but it provides an intuition
-  to how these systems function.
+  This brief list includes some lossy methods, though this
+  presentation will cover only lossless methods.
 \end{frame}
@@
   $c : \Sigma^n \to \bigcup_{k = 1}^{n} \Sigma^k$ for all
-  $n \in \mathbb{N}$ such that $n \geq 2$. A compression scheme is
-  considered ``perfect'' if and only if its image is the set of all
-  strings of length strictly less than that of the domain.
+  $n \in \mathbb{N}$ such that $n \geq 2$.
 \end{frame}
@@
   \subsection{Impossibility of perfect compression}
+  A compression scheme is considered ``perfect'' if and only if its
+  image is the set of all strings of length strictly less than that of
+  the domain.
   \begin{Theorem}
@@
 \begin{frame}
+  \subsection{Maximum Compressibility of Data}
+  \begin{itemize}
+  \item This result suggests there is a limit to how small compressed
+    data can be.
+  \item Shannon's Noiseless Coding gives a lower bound.
+    \begin{itemize}
+    \item The average word size of a prefix code on a channel is
+      greater than or equal to its entropy.
+    \item Since human data sources are typically very regular
+      (i.e. have low information entropy), data compression methods
+      are able to greatly reduce the sizes of these sorts of data.
+    \end{itemize}
+  \end{itemize}
+\end{frame}
 
+\begin{frame}
+  \section{Lempel-Ziv Based Algorithms}
+  \subsection{LZ}
+  \frametitle{Lempel-Ziv compression algorithms}
+  LZ77 and LZ78 (a.k.a. LZ1 and LZ2, a.k.a. LZ)
+  \begin{itemize}
+  \item Introduced in 1977 and 1978 by Abraham Lempel and Jakob Ziv
+  \item Dictionary block compressor family
+  \item Uses a sliding-window approach
+  \item Produces 3-tuples representing prefixes
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \subsection{LZ77 Example}
+  Suppose we are to encode the tongue-twister ``Can you can a can as a
+  canner can can a can?'' with a window size of 8.
+\end{frame}
+
+\begin{frame}
+  \section{FLAC}
+  \frametitle{Free Lossless Audio Codec}
+  The Free Lossless Audio Codec is a domain-specific compression
+  scheme for hi-fi audio.
 \end{frame}

### 2018-12-10 paper.tex
Update the paper and presentation.

@@
 \author{Jack Stanek}
+\date{December 10, 2018}
 
@@
 illustrate the implications of this result with a simple
-example. Suppose we are attempting to compress a particular channel
-$X$ which is text document in which the frequencies of every letter of
-the Latin alphabet are distributed equally. (It is extremely unlikely
-that a text in any human language would have this property; it is used
-here to illustrate a pathological edge case in general.) Then the
-average word length must be greater than
+example. Suppose we are attempting to compress some arbitrary channel,
+$X$, which is text document in which the frequencies of every letter
+of the Latin alphabet are distributed equally. (It is extremely
+unlikely that a text in any human language would have this property;
+it is used here to illustrate a pathological edge case in general.)
+Then the average word length must be greater than
 $26 * -\frac{1}{26} \log_2{\frac{1}{26}} \approx 4.7$ bits. If we were
@@
 
-To illustrate: suppose we are to encode the string ``oodles of
-noodles'' using LZ77, with a window of length 8. The string is 16
-characters long, requiring 16 bytes to store. The current character is
-set to ``o'', the first character of the string. However, since there
-are two ``o'' characters in a row, the algorithm can ``overshoot'' the
-end of the window in order to efficiently implement run-length
-encoding (that is, recording a character and the number of copies
-folloiwng it.) Thus, \texttt{(2, 0, 'o')} is emitted, and the current
-character is incremented by two. The algorithm then emits
-single-character prefix 3-tuples for \texttt{d, l, e, s, ' ',} (the
-space character), \texttt{o}, and \texttt{f}. At
+To illustrate: suppose we want to encode the laughing string
+``hahahahaha'' using the LZ77 algorithm with an 8-character window. We
+would begin by encoding the first two (unique) characters as
+\texttt{(0, 0, 'h')} and \texttt{(0, 0, 'a')}, respectively. Then, the
+rest of the word can be encoded as \texttt{(2, 8, '\textbackslash
+  0')}, where \texttt{'\textbackslash 0'} represents the ASCII null
+character. The resultant output can be stored in 9 bytes, compared to
+the 11 required by the original input string. This example also
+illustrates how the algorithm can use a powerful implementation or
+run-length encoding; LZ77 can represent prefixes which extend past the
+end of the sliding window by ``feeding'' output characters back into
+the decoder, thus allowing repetative cyclic prefixes to be
+represented compactly.
+
+% To illustrate: suppose we are to encode the tongue-twister ``can you
+% can a can as a can man can can a can'' using LZ77, with a window of
+% length 8. The string is 44 characters long, thus requiring 44 bytes to
+% store. In its implementation, the algorithm implicitly holds a
+% windows-worth full of null-character bytes (represented in text as
+% \texttt{\textbackslash 0 or 0x00}) sets the current character to the
+% first character of the input. Since, in this case, the first 7
+% characters of the input are all unique, and as such they cannot be
+% represented by duplicate prefixes from earlier in the text (as there
+% is no ``earlier text'' which matches it.) Thus each of these first
+% bytes is emitted as \texttt{(0, 1, character)}. By this point, the
+% current prefix has advanced to the second space character. Since there
+% is a space less than 8 characters before this, the next character
+% (\texttt{'c'}) is appended to the current prefix. Since the prefix
+% \texttt{' c'} is not found in the window, the algorithm emits
+% \texttt{(4, 1, 'c')}. The next prefix is set to \texttt{'a'} and the
+% process repeats. This time, the prefix \texttt{'an '} can be found in
+% the window, so the algorithm proceeds up to the following \texttt{'c'}
+% and then emits \texttt{(7, 3, 'c')}. The next prefix is set to
+% \texttt{'a'} and the algorithm continues. For brevity, the remaining
+% steps will not be listed, but the compressed version contains 17
+% 3-tuples, requiring 53 bytes to store. This example shows not only
+% that any compression algorithm has incompressable inputs, but also
+% that the design of an algorithm can affect which inputs are
+% incompressable. (In this case, the first few bytes are encoded as
+% three-byte triples, whereas later bytes are encoded as pure
+% prefixes. Since this example is short, the cost incurred by the first
+% bytes is too high to be overcome by the later savings.)
 
@@
 worthwhile to study some compression scheme which is specifically
-designed to store a particular type of data. In this secion I will
+designed to store a particular type of data. In this section I will
 describe the structure and function of the Free Lossless Audio Codec,

### 2018-12-10 presentation/presentation.tex
Update the paper and presentation.

@@
 \begin{frame}
-  \subsection{Impossibility of perfect compression}
+  \subsection{Impossibility of Perfect Compression}
   A compression scheme is considered ``perfect'' if and only if its
@@
   \section{Lempel-Ziv Based Algorithms}
-  \subsection{LZ}
+  \subsection{LZ77 and LZ78}
   \frametitle{Lempel-Ziv compression algorithms}
@@
   \subsection{LZ77 Example}
-  Suppose we are to encode the tongue-twister ``Can you can a can as a
-  canner can can a can?'' with a window size of 8.
+  \frametitle{A Brief Example}
+
+  Suppose we are to encode the word ``hahahahaha'' with a window size
+  of 16.
+\end{frame}
+
+\begin{frame}
+  \frametitle{LZ78}
+  LZ78 is a slightly different modification on the LZ77 algorithm.
+  \begin{itemize}
+  \item Instead of a sliding window, it uses a ``chained dictionary''
+  \item Each entry in the dictionary is a two-tuple: \texttt{(
+      previous prefix index, following character)}
+  \item The dictionary can be represented as a tree data structure
+  \item Equivalent to LZ77 even though it uses a different method
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \subsection{LZW and DEFLATE}
+  \frametitle{Derivative Methods: LZW and DEFLATE}
+
+  Since they are very effective and efficient algorithms, LZ77 and
+  LZ78 have inspired a wide variety of derivative methods.
+
+  \begin{itemize}
+  \item LZW (Lempel-Ziv-Welch): A method based on LZ78 which begins
+    with a pre-populated prefix dictionary, with each individual
+    character in the alphabet.
+  \item DEFLATE: Splits the input into blocks, then encodes the
+    alphabet using a Huffman code, then compresses the resultant input
+    channel using LZ78.
+  \end{itemize}
 \end{frame}
@@
   scheme for hi-fi audio.
+  \begin{enumerate}
+  \item Split input into blocks (can optimize for compression or
+    speed)
+  \item Encode stereo channels (left-right or mid-side)
+  \item Fit the sound signal to an n-degree polynomial (configurable)
+  \item Calculate the residuals (difference between the raw signal and
+    the fit polynomial)
+  \item Encode the residual values using Rice codes (essentially a
+    Huffman coding optimized for alphabets with geometric
+  \end{enumerate}
 \end{frame}
 
+\begin{frame}
+  \section{Question and Answer}
+  \frametitle{Any questions?}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Thank you!}
+\end{frame}
 \end{document}
